---
title: "Homework 02"
subtitle: "ECON 2200 — Fall 2025"
author: "Ben Kassan"
date: "September 13, 2025"
output: pdf_document
---

Q1a

What the study set out to do:

Bertrand and Mullainathan ran an experiment to test whether employers discriminate based on perceived race. They sent comparable résumés to real job ads in Boston and Chicago, randomly assigning first names that commonly signal either Black or White identities. They also randomized résumé quality (e.g., stronger vs. weaker credentials). The main outcome was whether the employer called the applicant back.

Why this identifies a causal effect:

Because the name signals and résumé quality were randomly assigned, the groups are on average—identical in every respect except the perceived race conveyed by the name. That means any systematic difference in callback rates can be attributed to race, not to unobserved applicant traits. In other words, the randomized design turns the question from correlational to causal. The difference in mean callback rates across name groups is an unbiased estimate of the average causal effect of perceived race on employer interest. The quality randomization also lets them test whether stronger credentials help applicants with Black-sounding names as much as those with White-sounding names. The causal effect is reinforced with a large sample size of 4,870.

Q1b

```{r}
# Q1(b) Simple Script
# Economics Statistics Class

# Load data
resume <- read.csv("/Users/benkassan/Desktop/rdata/ResumeNames.csv", stringsAsFactors = FALSE)

# Recode race from ethnicity
eth <- tolower(resume$ethnicity)
race <- ifelse(eth %in% c("afam","black"), "Black-sounding",
         ifelse(eth %in% c("cauc","white"), "White-sounding", NA))

# Recode callback from call
cal <- tolower(as.character(resume$call))
callback <- ifelse(cal %in% c("yes","y","1","true"), 1,
            ifelse(cal %in% c("no","n","0","false"), 0, NA))

# Keep only rows with race info
keep <- !is.na(race)
race <- race[keep]
callback <- callback[keep]

# Calculate totals and rates
n <- tapply(callback, race, function(x) sum(!is.na(x)))
cb <- tapply(callback, race, function(x) sum(x == 1, na.rm = TRUE))
rate <- cb / n

# Print results
results <- data.frame(
  race = names(n),
  n = as.integer(n),
  callbacks = as.integer(cb),
  callback_rate = round(100 * rate, 1),
  row.names = NULL
)

print(results)

```

```{r message=FALSE}
# Q1(c) Graphs by Gender and Race

# Load data
resume <- read.csv("/Users/benkassan/Desktop/rdata/ResumeNames.csv", stringsAsFactors = FALSE)

# Recode race
eth <- tolower(resume$ethnicity)
resume$race <- ifelse(eth %in% c("afam","black"), "Black-sounding",
                 ifelse(eth %in% c("cauc","white"), "White-sounding", NA))

# Recode callback
cal <- tolower(as.character(resume$call))
resume$callback <- ifelse(cal %in% c("yes","y","1","true"), 1,
                    ifelse(cal %in% c("no","n","0","false"), 0, NA))

# Keep rows with race + gender
resume <- subset(resume, !is.na(race) & !is.na(gender))

# Summarize by race and gender
library(dplyr)
by_group <- resume %>%
  group_by(race, gender) %>%
  summarize(rate = mean(callback, na.rm = TRUE), .groups = "drop")

# Plot: grouped bar chart
library(ggplot2)
ggplot(by_group, aes(x = race, y = rate, fill = gender)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Perceived Race", y = "Callback Rate",
       title = "Callback Rates by Race and Gender",
       fill = "Gender") +
  theme_minimal()

```

In our sample, men received fewer callbacks than women in both perceived race groups. For White-sounding names, women’s callback rate is higher than men’s (9.9% vs. 8.9%); for Black-sounding names, the same pattern holds (6.6% vs. 5.8%). In short, being male is associated with a lower chance of a callback across both groups. The White–Black gap is 3.3 percentage points for women and 3.1 points for men—very similar. Given how small that 0.2-point difference is, I wouldn’t claim a larger discrimination effect for women without a significance test; it’s likely just noise.

2a

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Load and clean the data
df <- read.csv("/Users/benkassan/Desktop/rdata/ResumeNames.csv", stringsAsFactors = FALSE) %>%
  mutate(
    # Standardize race coding
    race = case_when(
      tolower(ethnicity) %in% c("afam", "black", "b") ~ "African-American",
      tolower(ethnicity) %in% c("cauc", "white", "w") ~ "Caucasian",
      TRUE ~ NA_character_
    ),
    # Standardize gender coding
    gender = case_when(
      tolower(gender) %in% c("male", "m") ~ "Male",
      tolower(gender) %in% c("female", "f") ~ "Female",
      TRUE ~ NA_character_
    ),
    # Convert callback to numeric (handle various formats)
    callback = case_when(
      tolower(as.character(call)) %in% c("yes", "y", "1", "true") ~ 1,
      tolower(as.character(call)) %in% c("no", "n", "0", "false") ~ 0,
      TRUE ~ NA_real_
    ),
    # Standardize city names
    city = tolower(city)
  ) %>%
  filter(!is.na(race), !is.na(gender), !is.na(city), !is.na(callback))

# Calculate callback rates by city, gender, and race
summary_data <- df %>%
  group_by(city, gender, race) %>%
  summarize(
    callback_rate = mean(callback, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Calculate the racial difference (Caucasian - African-American)
racial_diff <- summary_data %>%
  pivot_wider(
    names_from = race,
    values_from = c(callback_rate, n),
    names_sep = "_"
  ) %>%
  mutate(
    # Calculate difference in percentage points
    diff_pp = (`callback_rate_Caucasian` - `callback_rate_African-American`) * 100,
    # Create labels
    diff_label = sprintf("%+.1f pp", diff_pp),
    # Create sample size labels
    n_label_white = sprintf("nW=%d", n_Caucasian),
    n_label_black = sprintf("nB=%d", `n_African-American`)
  ) %>%
  # Order factors for proper display
  mutate(
    gender = factor(gender, levels = c("Female", "Male")),
    city = factor(city, levels = c("boston", "chicago"))
  )

# Create the plot
p <- ggplot(racial_diff, aes(x = gender, y = diff_pp, fill = gender)) +
  # Add horizontal line at 0
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", alpha = 0.5) +
  
  # Create bars
  geom_col(width = 0.7, show.legend = FALSE) +
  
  # Add difference labels above bars
  geom_text(aes(label = diff_label), 
            vjust = -0.5, 
            size = 5, 
            fontface = "bold") +
  
  # Add sample size for White names
  geom_text(aes(label = n_label_white, y = diff_pp * 0.3), 
            hjust = 0.5,
            size = 3.5, 
            color = "grey30") +
  
  # Add sample size for Black names  
  geom_text(aes(label = n_label_black, y = diff_pp * 0.5), 
            hjust = 0.5,
            size = 3.5, 
            color = "grey30") +
  
  # Facet by city
  facet_wrap(~ city, nrow = 1, scales = "free_x") +
  
  # Custom colors matching the reference image
  scale_fill_manual(values = c("Female" = "#FF6B6B", "Male" = "#4ECDC4")) +
  
  # Set y-axis limits
  scale_y_continuous(
    limits = c(-1, 6),
    breaks = seq(0, 5, 1),
    expand = c(0, 0)
  ) +
  
  # Labels
  labs(
    title = "Racial Difference in Callback Rates by City × Gender",
    subtitle = "Positive values indicate higher callback rates for Caucasian names",
    x = NULL,
    y = "Difference in callback rate (percentage points)"
  ) +
  
  # Clean theme
  theme_minimal(base_size = 14) +
  theme(
    # Title styling
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 12, hjust = 0, color = "grey30"),
    
    # Facet styling
    strip.text = element_text(face = "bold", size = 14),
    strip.background = element_rect(fill = "grey95", color = NA),
    
    # Axis styling
    axis.text.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 11),
    axis.title.y = element_text(size = 12, margin = margin(r = 10)),
    
    # Grid styling
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey90"),
    
    # Panel spacing
    panel.spacing = unit(1.5, "lines"),
    
    # Overall plot margins
    plot.margin = margin(10, 10, 10, 10)
  )

# Display the plot
print(p)

```

2b

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)

# Load and clean the data
df <- read.csv("/Users/benkassan/Desktop/rdata/ResumeNames.csv", stringsAsFactors = FALSE) %>%
  mutate(
    # Standardize race coding
    race = case_when(
      tolower(ethnicity) %in% c("afam", "black", "b") ~ "African-American",
      tolower(ethnicity) %in% c("cauc", "white", "w") ~ "Caucasian",
      TRUE ~ NA_character_
    ),
    # Convert callback to numeric
    callback = case_when(
      tolower(as.character(call)) %in% c("yes", "y", "1", "true") ~ 1,
      tolower(as.character(call)) %in% c("no", "n", "0", "false") ~ 0,
      TRUE ~ NA_real_
    ),
    # Standardize city names for display
    city_clean = case_when(
      tolower(city) == "boston" ~ "Boston",
      tolower(city) == "chicago" ~ "Chicago",
      TRUE ~ tools::toTitleCase(tolower(city))
    )
  ) %>%
  filter(!is.na(race), !is.na(city), !is.na(callback), !is.na(jobs))

# Calculate mean callback rate for EACH combination of jobs, race, and city
# This properly aggregates across ALL applicants (both men and women)
summary_jobs <- df %>%
  group_by(jobs, race, city_clean) %>%
  summarize(
    callback_rate = mean(callback, na.rm = TRUE) * 100,  # Convert to percentage
    n = n(),
    .groups = "drop"
  )

# Create the plot faceted by race and city as requested
p <- ggplot(summary_jobs, aes(x = jobs, y = callback_rate)) +
  
  # Add line connecting points
  geom_line(color = "#2C3E50", linewidth = 1.2, alpha = 0.9) +
  
  # Add points with size based on sample size
  geom_point(aes(size = n), color = "#2C3E50", alpha = 0.9) +
  
  # Facet by race and city (creates 4 panels: 2 races × 2 cities)
  facet_grid(race ~ city_clean) +
  
  # Size scale for sample sizes
  scale_size_continuous(
    range = c(2, 10),
    breaks = c(100, 300, 500, 700, 900),
    name = "Sample Size"
  ) +
  
  # X-axis settings
  scale_x_continuous(
    breaks = seq(1, 7, 1),
    limits = c(0.8, 7.2)
  ) +
  
  # Y-axis settings
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    limits = c(0, NA),
    expand = expansion(mult = c(0.02, 0.08))
  ) +
  
  # Labels
  labs(
    title = "Impact of Number of Jobs on Callback Rates by Race and City",
    subtitle = "Mean callback rates aggregated across all applicants",
    x = "Number of Jobs Listed on Resume",
    y = "Callback Rate",
    caption = "Each panel shows a race-city combination. Point size indicates sample size."
  ) +
  
  # Clean theme
  theme_bw(base_size = 12) +
  theme(
    # Title and text
    plot.title = element_text(face = "bold", size = 16, hjust = 0),
    plot.subtitle = element_text(size = 12, hjust = 0, color = "grey30"),
    plot.caption = element_text(size = 9, hjust = 0, color = "grey50", margin = margin(t = 10)),
    
    # Facet styling
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "grey90"),
    
    # Legend
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    legend.box.margin = margin(t = -5),
    
    # Grid
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey85", linetype = "dotted"),
    
    # Spacing
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(10, 10, 10, 10),
    
    # Axis text
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )

# Print the plot
print(p)
```


3a:

We want to show that for any real numbers $a, b$,
\[
\frac{1}{n} \sum_{i=1}^n (a x_i + b) \;=\; a \bar{x} + b,
\]
where the sample mean is defined as
\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i.
\]

---

**Proof.**  
We begin with the left-hand side:
\[
\frac{1}{n} \sum_{i=1}^n (a x_i + b).
\]

Distributing the summation across terms gives:
\[
\frac{1}{n} \left( \sum_{i=1}^n a x_i + \sum_{i=1}^n b \right).
\]

Next, we can pull the constant $a$ outside of the first summation, and recognize that $\sum_{i=1}^n b = n b$:
\[
\frac{1}{n} \left( a \sum_{i=1}^n x_i + n b \right).
\]

Breaking this apart, we get:
\[
\frac{a}{n} \sum_{i=1}^n x_i \;+\; \frac{n b}{n}.
\]

Simplifying, the first term is $a \bar{x}$ (by the definition of the sample mean), and the second term is simply $b$:
\[
a \bar{x} + b.
\]

Thus, we have shown:
\[
\frac{1}{n} \sum_{i=1}^n (a x_i + b) \;=\; a \bar{x} + b.
\]
\[
\boxed{\text{Q.E.D.}}
\]


3b

We want to show that
\[
\frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^2 
= \frac{1}{n}\sum_{i=1}^n x_i^2 - (\bar{x})^2.
\]

---

**Proof.**  
We begin with the left-hand side:
\[
\frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^2.
\]

Expanding the square gives:
\[
\frac{1}{n} \sum_{i=1}^n \left( x_i^2 - 2x_i\bar{x} + \bar{x}^2 \right).
\]

Separating terms inside the summation:
\[
\frac{1}{n} \left( \sum_{i=1}^n x_i^2 - 2\bar{x}\sum_{i=1}^n x_i + \sum_{i=1}^n \bar{x}^2 \right).
\]

Since $\sum_{i=1}^n x_i = n\bar{x}$ and $\sum_{i=1}^n \bar{x}^2 = n\bar{x}^2$, we obtain:
\[
\frac{1}{n} \left( \sum_{i=1}^n x_i^2 - 2n\bar{x}^2 + n\bar{x}^2 \right).
\]

Simplifying:
\[
\frac{1}{n} \left( \sum_{i=1}^n x_i^2 - n\bar{x}^2 \right).
\]

Breaking apart the fraction yields:
\[
\frac{1}{n}\sum_{i=1}^n x_i^2 - \bar{x}^2.
\]

Thus, we have shown:
\[
\frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^2 
= \frac{1}{n}\sum_{i=1}^n x_i^2 - (\bar{x})^2.
\]
\[
\boxed{\text{Q.E.D.}}
\]
